\documentclass[a4paper,12pt]{article}
\include{../preamble}
% The title of the report
\title{Literature Review: Block Identifiability and Misalignment in Multimodal Contrastive Learning}
% Write your name in author
\author{Dowoo Kim, Ryan Jackson, David Zhao}

\begin{document}
\maketitle

Discovering latent structures through sensor measurements remains an important problem in machine learning. Many algorithms have been proposed to identify these latent structures, such as linear independent component analysis (ICA) \cite{Comon1994} \cite{HyvarinenOja2000}, non-linear ICA \cite{pmlr-v89-hyvarinen19a}, and even some deep learning-based approaches \cite{Bengio2013}. However, these methods often rely on strict assumptions about the latent factors, such as independence or non-Gaussianity, which are too restrictive for practical purposes. Recent work in causal representation learning \cite{scholkopf2021causalrepresentationlearning} began relaxing the independence assumption to allow for some form of dependence between latent factors, though these methods still remain too limiting in their assumptions at large. \\

To that end, von Kügelgen et al. (2021) introduced the notion of \textit{block-identifiability} \cite{von2021self}. In their paper, they model augmented views using a latent-variable framework consisting of content and style components. In this framework, content remains unchanged across views, whereas style varies between views. The style of an augmented view is assumed to be generated by first randomly selecting a subset of the original view’s style variables and then replacing their values with samples drawn from a distribution conditioned on the values of that subset in the original style. The authors show that, under mild assumptions–allowing for dependencies among latent variables–the true content partition is identifiable up to an invertible transformation of the inferred content block. They refer to this notion—identifying a set of latent variables only up to such a transformation—as \textit{block identification}.

This result departs from the traditional goal of disentangling individual, independent latent factors. Instead, it focuses on identifying blocks of variables that are preserved across augmentations. This perspective is more flexible and realistic, as it allows dependencies among latent factors within each block while still enabling the recovery of shared structure across views. Importantly, the authors show that such blocks can be identified by minimizing suitable contrastive objectives. In their most general setting, identification is achieved by minimizing an alignment objective with maximum-entropy regularization, which encourages representations to be both aligned across views and informative. The widely used InfoNCE loss can be understood as an asymptotic approximation of this objective, providing a direct link between block-identifiability and standard contrastive learning practice. Their paper also introduces the Causal3DIdent benchmark, which has become a standard dataset for studying identifiable representation learning with structured latent factors. \\

Following this foundational work, several studies have extended block-identifiability to multimodal settings, particularly focusing on image–caption pairs. Of special relevance is Daunhawer et al (2023), who prove that contrastive learning can block-identify content factors shared between two modalities. Their latent-variable framework, like that of von Kügelgen et al. (2021), consists of invariant content and variable style components, but further incorporates modality-specific components that capture factors unique to each modality \cite{daunhawer2023identifiabilityresultsmultimodalcontrastive}. \\

An important consideration in the multimodal setting is the potential misalignment between modalities—for example, an image and its caption may not perfectly reflect the same underlying semantic information. Cai et al. (2025) reformulate the framework of Daunhawer et al. (2023) to explicitly model such misalignment by introducing two mechanisms: selection bias and perturbation bias  \cite{cai2025valuecrossmodalmisalignmentmultimodal}. In their framework, the latent variable space is divided into shared semantic factors, which underlie both modalities, and modality-specific components, which capture factors unique to each modality. Within the shared semantic space, misalignment is modeled by applying the biases to only one modality. Selection bias corresponds to semantic factors that are entirely missing in the affected modality, while perturbation bias corresponds to factors that are present but distorted or noisy in that modality. In this formulation, contrastive learning is able to identify the unbiased semantic factors. \\

Thoreau et al. (2025) investigate contrastive learning’s ability to capture modality-specific information from paired multimodal data in a simplified linear setting \cite{thoreau2025multimodalrepresentationlearningalignment}. They consider an objective consisting of two prediction losses—one for each modality, corresponding to modality-specific targets—together with an alignment term encouraging the representations of the two modalities to match, under the constraint that the prediction loss for the second modality is minimized. Under the assumptions that the observations of the second modality are sufficiently informative for predicting their own targets, but only weakly informative for predicting the targets of the first modality, they show that the resulting solution minimizes neither the first prediction loss nor the alignment loss. In particular, this reveals a fundamental trade-off between the two downstream prediction tasks, which serve as proxies for modality-specific information. Viewed through the framework of Cai et al. (2025), such asymmetric informativeness can arise from selection bias, where semantic variables missing from one modality reduce its ability to predict targets associated with the other modality. From this perspective, the trade-off identified by Thoreau et al. (2025) illustrates how selection bias can lead to fundamental limitations in alignment-based multimodal representation learning. \\

Finally, Yang et al. (2024) study multimodal robustness in the context of downstream classification \cite{ICLR2024_45a9d429}. They introduce the notion of modality preference, which describes the extent to which a model’s predictions rely on each modality, and analyze robustness to perturbations applied to either modality. In particular, they define a perturbation radius, measuring the magnitude of corruption that can be tolerated before the classifier’s prediction changes, and derive a lower bound on this radius in terms of integration factors related to modality preference. When semantic variables are categorical, recovering them from a learned representation can be viewed as a classification problem; under this interpretation, the perturbation radius quantifies how much distortion of the observations can be tolerated before the inferred semantic label changes. Viewed through the framework of Cai et al. (2025), these distortions correspond to a more general form of perturbation bias, allowing shared semantic factors to be corrupted in either modality.


\newpage
\bibliographystyle{unsrt}
\bibliography{ref.bib}
\end{document}
